### 2021-02-23 paper1/paper1.tex
Lots of improvements

% \author[1]{Thomas Kehrenberg}
% \author[1,2]{Zexun Chen}
% \author[1]{Novi Quadrianto}
% \affil[1]{Predictive Analytics Lab (PAL), University of Sussex, Brighton, UK}
% \affil[2]{BioComplex Laboratory, University of Exeter, Exeter, UK}
% \title{Tuning Fairness by Balancing Target Labels}

@@

% ============= content ====================
\section{Abstract}

@@

\section{Target labels for tuning group fairness}
We will start by describing several notions of group fairness.
For each individual, we have a vector of non-sensitive attributes $x\in\mathcal{X}$, a class label $y\in\mathcal{Y}$, and a sensitive attribute $s\in\mathcal{S}$ (e.g.\ racial origin or gender).
%
We focus on the case where $s$ and $y$ are binary. 
%
We assume that a positive label $y=1$ corresponds to a positive outcome for an individual -- for example, being accepted for a loan.
%
%
\emph{Group fairness} balances a certain condition between groups of individuals with different sensitive attribute, $s$ versus $s\sp{\prime}$. 
%
The term $\hat{y}$ below is the prediction of a machine learning model that,
in most works, uses only non-sensitive attributes $x$.
%
Several group fairness criteria have been proposed (e.g. \citet{zafar2017fairnesstreatment,chouldechova2017fair,hardt2016equality}):
\begin{align}
& \text{equality of positive rate (Demographic Parity):}\nonumber\\
& \text{Pr}(\hat{y}=1|s)=\text{Pr}(\hat{y}=1|s\sp{\prime}) \label{eq:eq_dem_par}\\
& \text{equality of accuracy:}\nonumber\\
& \text{Pr}(\hat{y}=y|s)=\text{Pr}(\hat{y}=y|s\sp{\prime}) \\
& \text{equality of true positive rate (Equality of Opportunity):} \nonumber\\
& \text{Pr}(\hat{y}=1|s,y=1)=\text{Pr}(\hat{y}=1|s\sp{\prime},y=1)  \label{eq:eq_opp_cri}~.
\end{align}
\emph{Equalised odds} criterion corresponds to Equality of Opportunity (\ref{eq:eq_opp_cri}) plus equality of false positive rate.

@@

\subsection{Choice of parameters}\label{sec:dp}
A balanced dataset is characterised by an independence of the label $\bar{y}$ and the sensitive attribute $s$.
%As our model is predicting $\bar{y}$, we want this independence to hold for $\bar{y}$.
%The graphical model in Fig~\ref{fig:biasfeaturebiaslabel} does not imply $\bar{y} \bot s$,
Given that we have complete control over the \emph{transition probabilities},
we can ensure this independence by requiring $P(\bar{y}=1|s=0)=P(\bar{y}=1|s=1)$.
Our constraint is then that both of these probabilities are equal to the same value,
which we will call the target rate $\mathit{PR}_t$ (``PR'' as \emph{positive rate}):
\begin{align}
  &P(\bar{y}=1|s=0) \overset{!}{=} \mathit{PR}_t\quad\text{and}%\nonumber\\
  \quad P(\bar{y}=1|s=1) \overset{!}{=} \mathit{PR}_t~.
\end{align}
This leads us to the following constraints for $s\prime\in\{0, 1\}$:
\begin{align}
  \mathit{PR}_t &= P(\bar{y}=1|s=s\prime) %\nonumber\\
  =\sum\limits_y P(\bar{y}=1|y,s=s\prime)\, P(y|s=s\prime).\label{eq:dpconstraint}
\end{align}
%$P(y|s=s\prime)$ has to be set to the value at which we want our constraint to hold.
We call $P(y=1|s=j)$ the base rate $\mathit{PR}_b^j$ which we estimate from the training set:
\begin{align*}
  P(y=1|s=i) = \frac{\text{number of points with } y=1 \text{ in group }i}
  {\text{number of points in group }i}~.
\end{align*}
Expanding the sum, we get
\begin{align}
  &\mathit{PR}_t \nonumber\\
  =\,\, &P(\bar{y}=1|y=0,s=s\prime) \cdot (1-\mathit{PR}_b^1) 
  +P(\bar{y}=1|y=1,s=s\prime) \cdot \mathit{PR}_b^1~.\label{eq:dpconstexpanded}
\end{align}
This is a system of linear equations consisting of two equations (one for each value of $s\prime$)
and four free variables: $P(\bar{y}=1|y,s)$ with $y,s\in\{0, 1\}$.
The two unconstrained degrees of freedom determine how strongly the accuracy will be affected by the fairness constraint.
If we set $P(\bar{y}=1|y=1,s)$ to 0.5,
then this expresses the fact that a train label $y$ of $1$ only implies a target label $\bar{y}$ of $1$ in 50\% of the cases.
In order to minimise the effect on accuracy,
we make $P(\bar{y}=1|y=1,s)$ as high as possible and $P(\bar{y}=1|y=0,s)$, conversely, as low as possible.
However, the lowest and highest possible values are not always 0 and 1 respectively.
To see this, we solve for $P(\bar{y}=1|y=0,s=j)$ in \eqref{eq:dpconstexpanded}:
\begin{align}
  &P(\bar{y}=1|y=0,s=j) %\nonumber\\
  =\,\, \frac{\mathit{PR}_b^j}{1-\mathit{PR}_b^j} \left(\frac{\mathit{PR}_t}{\mathit{PR}_b^j} - P(\bar{y}=1|y=1,s=j)\right)~.
\end{align}
If $\nicefrac{\mathit{PR}_t}{\mathit{PR}_b^j}$ were greater than 1,
then setting $P(\bar{y}=1|y=0,s=j)$ to 0
would imply a $P(\bar{y}=1|y=1,s=j)$ value greater than 1.
A visualisation that shows why this happens can be found in the Supplementary Material.
We thus arrive at the following definitions:
\begin{align}
  P(\bar{y}=1|y=1,s=j)&=\begin{cases}
    1 &\text{if }\mathit{PR}_t>\mathit{PR}_b^j\\
    \frac{\mathit{PR}_t}{\mathit{PR}_b^j} &\text{otherwise.}
  \end{cases}%
  \label{eq:tau-11}\\
  P(\bar{y}=1|y=0,s=j)&=\begin{cases}
    \frac{\mathit{PR}_t-\mathit{PR}_b^j}{1-\mathit{PR}_b^j} &\text{if }\mathit{PR}_t>\mathit{PR}_b^j\\
    0 &\text{otherwise.}
  \end{cases}%
  \label{eq:tau-01}
\end{align}
Algorithm~\ref{alg:parity} shows pseudocode of the procedure, including the computation of the allowed minimal and maximal value.

@@

\subsection{Conditionally balanced dataset}
\noindent There is a fairness definition related to demographic parity which allows conditioning on ``legitimate'' risk factors $\ell$
when considering how equal the demographic groups are treated~\cite{corbett2017algorithmic}.
This cleanly translates into balanced datasets which are balanced conditioned on $\ell$:
\begin{align}
  &P(\bar{y}=1|\ell=\ell\prime, s=0) \overset{!}{=} P(\bar{y}=1|\ell=\ell\prime, s=1)~.
\end{align}
We can interpret this as splitting the data into partitions based on the value of $\ell$,
where the goal is to have all these partitions be balanced.
This can easily be achieved by our method by setting a $\mathit{PR}_t(\ell)$ for each value of $\ell$
and computing the transition probabilities for each sample depending on $\ell$.

\section{Related work}\label{sec:relatedwork}
There are several ways to enforce fairness in machine learning models:
as a pre-processing step
% (e.g.~),
\citep{kamiran2012data,louizos2015variational,lum2016statistical,zemel2013learning,Chiappa19,QuaShaTho19},
% kusner2017counterfactual
as a post-processing step \citep{feldman2015certifying,hardt2016equality}, %(e.g.~\citep{feldman2015certifying,hardt2016equality}),
 or as a constraint during the learning phase
% (e.g.~).
\citep{calders2009building,zafar2017fairnesstreatment,zafar2017fairnessconstraints,DonOneBenShaetal18,DimLiuParRad19}.
%
Our method enforces fairness during the learning phase (an in-processing approach) but, unlike other approaches, we do not cast fair-learning as a \emph{constrained} optimisation problem.
Constrained optimisation requires a customised procedure.
%
In \citet{GohCotGupFri16}, \citet{zafar2017fairnesstreatment}, and \citet{zafar2017fairnessconstraints},
% In~\citet{GohCotGupFri16,zafar2017fairnesstreatment,zafar2017fairnessconstraints},
suitable majorisation-minimisation/convex-concave procedures~\citep{LanSri09} were derived.
%
Furthermore, such constrained optimisation approaches may lead to more unstable training,
and often yield classifiers with both worse accuracy and more unfair~\citep{CotJiaWanNar18}.

@@

The massaging approach uses a classifier to first rank all samples according to their probability of having a positive label ($y=1$)
and then flips the labels that are closest to the decision boundary such that the data then satisfies demographic parity.
This \emph{pre-processing} approach is similar in spirit to our \emph{in-processing} method but differs in the execution.
In our method (Section \ref{sec:dp}), ``ranking'' and classification happen in one step and labels are not explicitly flipped
but assigned probabilities of being flipped.
%{\color{red}This protects us from flipping the wrong labels.}

@@

In addition to the GP and LR baselines, we compare our proposed model with the following methods:
Support Vector Machine (\emph{SVM}), \emph{Kamiran \& Calders}~\citep{kamiran2012data} (``reweighing'' method),
\emph{Agarwal et al.}~\citep{AgaBeyDudLanetal18} (using logistic regression as the classifier)
% Logistic Regression (\emph{LR})
and several methods given by Zafar et al.~\citep{zafar2017fairnessconstraints,zafar2017fairnesstreatment},
%\citet{zafar2017fairnessconstraints} and \citet{zafar2017fairnesstreatment},
which include maximising accuracy under demographic parity fairness constraints (\emph{ZafarFairness}),
maximising demographic parity fairness under accuracy constraints (\emph{ZafarAccuracy}),
and removing disparate mistreatment by constraining the false negative rate (\emph{ZafarEqOpp}).
% We make use of the fairness comparison by \citet{friedler2018comparative} %\citet{friedler2018comparative},
Every method is evaluated over 10 repeats that each have different splits of the training and test set.

\subsection{Results for Demographic Parity on Adult dataset}\label{sssec:demparresults}
\begin{figure}[t]
  \begin{center}
    \centerline{\includegraphics[width=0.98\columnwidth]{./figures/adult_tar_choice.eps}}
    \caption{Accuracy and fairness (demographic parity) for various target choices.
      (a): Adult dataset using race as the sensitive attribute;
      (b): Adult dataset using gender.
      Centre of the cross is the mean; height and width of the box encode half of standard derivation of accuracy and disparate impact.
    }%
    \label{fig:tar}
  \end{center}
  % \vspace*{-0.8cm}
% \caption{Accuracy and fairness for various targets.  w.r.t. accuracy and width of the box encodes standard derivation w.r.t. fairness criterion}
\end{figure}%
%
Following \citet{zafar2017fairnessconstraints} we evaluate demographic parity on the Adult dataset.
Table~\ref{tab:dempar} shows the accuracy and fairness for several algorithms.
In the table, and in the following, we use $\mathit{PR}_{s=i}$ to denote
the observed rate of positive predictions per demographic group $P (\hat{y} = 1 | s =i)$.
Thus, $\mathit{PR}_{s=0} / \mathit{PR}_{s=1}$ is a measure for demographic parity,
where a completely fair model would attain a value of $1.0$.
This measure for demographic parity is also called ``disparate impact''
(see e.g.~\citet{feldman2015certifying,zafar2017fairnesstreatment}).
%
As the results in Table~\ref{tab:dempar} show, FairGP and FairLR are clearly fairer than the baseline GP and LR\@.
We use the mean ($\mathit{PR}_t^{avg}$) for the target acceptance rate.
The difference between fair models and unconstrained models is not as large with \emph{race} as the sensitive attribute,
as the unconstrained models are already quite fair there.
The results of FairGP are characterised by high fairness and high accuracy.
FairLR achieves similar results to FairGP, but with generally slightly lower accuracy but better fairness.
We used the two step procedure of \citet{DonOneBenShaetal18}
to verify that we cannot achieve the same fairness result with just parameter search on LR\@.

In Fig.~\ref{fig:tar}, we investigate which choice of target ($\mathit{PR}_t^{avg}$, $\mathit{PR}_t^{min}$ or $\mathit{PR}_t^{max}$)
gives the best result.
We use $\mathit{PR}_t^{avg}$ for all following experiments as this is the fairest choice (cf. Section~\ref{sec:dp}).
The Fig.\ref{fig:tar}(a) shows results from Adult dataset with \emph{race} as sensitive attribute
where we have $\mathit{PR}_t^{min}=0.156$, $\mathit{PR}_t^{max}=0.267$ and $\mathit{PR}_t^{avg} =0.211$.
$\mathit{PR}_t^{avg}$ performs best in term of the trade-off.%equal (for \emph{race} as sensitive attribute)
%or better (for \emph{gender}) compared with the two other possibilities.
% The variant that uses $s$ for training and prediction (``use $s$'') performs significantly better here.
%\emph{ZafarAccuracy} can achieve good fairness results at the cost of accuracy.
\begin{figure}[t]
  \centering
  \includegraphics[width=0.98\textwidth]{./figures/adult_parity_scatter_pr_pr.eps}
  \caption{
    Predictions with different target acceptance rates (demographic parity) for 10 repeats.
    (a): $\mathit{PR}_{s=0}$ vs $\mathit{PR}_{s=1}$ using race as the sensitive attribute;
    (b): $\mathit{PR}_{s=0}$ vs $\mathit{PR}_{s=1}$ using gender.
  }%
  \label{fig:adult_parity_scatter_pr_pr}
\end{figure}
\begin{figure}[t]
  \centering
  \includegraphics[width=0.98\textwidth]{./figures/adult_parity_scatter_acc.eps}
  \caption{
    Predictions with different target acceptance rates (demographic parity) for 10 repeats.
    (a): disparate impact vs accuracy on Adult dataset using race as the sensitive attribute;
    (b): disparate impact vs accuracy using gender.
  }%
  \label{fig:adult_parity_scatter_acc}
\end{figure}

@@

\subsection{Results for Equality of Opportunity on ProPublica dataset.}\label{sssec:eqoppresults}
%
For equality of opportunity,
we again follow \citet{zafar2017fairnesstreatment} and evaluate the algorithm on the ProPublica dataset.
As we did for demographic parity,
we define a measure of equality of opportunity via the ratio of the true positive rates (TPRs) within the demographic groups.
We use $\mathit{TPR}_{s=i}$ to denote the observed TPR in group $i$: $P(\hat{y}=1|y=1, s=i)$,
and $\mathit{TNR}_{s=i}$ for the observed true negative rate (TNR) in the same manner.
The measure is then given by $\mathit{TPR}_{s=0} / \mathit{TPR}_{s=1}$.
A perfectly fair algorithm would achieve $1.0$ on the measure.
\begin{figure*}[t]
  \centering
  \includegraphics[width=0.98\textwidth]{./figures/propublica_opp_box_with_agarwal.eps}
%   \includegraphics[width=0.45\textwidth]{propublica-recidivism_race_opp_box.eps}
%   \includegraphics[width=0.45\textwidth]{propublica-recidivism_sex_opp_box.eps}
  \caption{Accuracy and fairness (with respect to \emph{equality of opportunity}) for various methods on ProPublica dataset.
    (a): using race as the sensitive attribute; (b): using gender.
    A completely fair model would achieve a value of 1.0 in the x-axis.
    %The best value for the TPR ratio is $1.0$.
    See Fig.~\ref{fig:propublica_opp_scatter_tpr}(a) and (b) on how these choices of PR setting translate to
    $\mathit{TPR}_{s=0}$ vs $\mathit{TPR}_{s=1}$. %and TPR vs TNR.
  }%
  \label{fig:propublica_opp_box}
\end{figure*}
\begin{figure*}[!ht]
  \centering
  \includegraphics[width=0.98\textwidth]{./figures/propublica_opp_scatter_tpr.eps}
  \caption{%
    Fairness measure $\mathit{TPR}_{s=0}$ vs $\mathit{TPR}_{s=1}$ (\emph{equality of opportunity}) for different target PRs ($\mathit{PR}_t$).
    (a): on dataset ProPublica recidivism using race as the sensitive attribute;
    (b): using gender.
  }%
  \label{fig:propublica_opp_scatter_tpr}
\end{figure*}

@@

\section{Discussion and conclusion}
Fairness is fundamentally not a challenge of algorithms alone, but very much a sociological challenge.
%
A lot of proposals have emerged recently for defining and obtaining fairness in machine learning-based decision making systems. 
%
% A vast amount
The vast majority of academic work has focused on two categories of definitions: statistical (group) notions of fairness and individual notions of fairness (see~\cite{verma2018fairness} for at least twenty different notions of fairness).
%
Statistical notions are easy to verify but do not provide protections to individuals. 
%
Individual notions do give individual protections but need strong assumptions, such as the availability of an agreed-upon similarity metric, which can be difficult in practice.
%
We acknowledge that a proper solution to algorithmic fairness cannot rely on statistics alone.
%
Nevertheless, these statistical fairness definitions can be helpful in understanding the problem and working towards solutions.
%
To facilitate this, at every step, the trade-offs that are present should be made very clear
and long-term effects have to be considered as well~\citep{liu2018delayed,kallus2018residual}.
% For example, it can be the case that false positives and false negatives are not equally costly,
% in which case a closer inspection of

>>>>

% \textsc{Authors}: Thomas Kehrenberg\footnote{}
% \author[1,2]{Zexun Chen}
% \author[1]{Novi Quadrianto}
% \affil[1]{Predictive Analytics Lab (PAL), University of Sussex, Brighton, UK}
% \affil[2]{BioComplex Laboratory, University of Exeter, Exeter, UK}
% \title{Tuning Fairness by Balancing Target Labels}

@@

% ============= content ====================
\chapter{Paper 1: Tuning Fairness by Balancing Target Labels}
\textsc{Authors}:\;\;Thomas Kehrenberg$^1$, Zexun Chen$^{1,2}$, and Novi Quadrianto$^1$ \\
\textsc{Affiliations}:\\
$^1$ Predictive Analytics Lab (PAL), University of Sussex, Brighton, UK\\
$^2$ BioComplex Laboratory, University of Exeter, Exeter, UK \\
\textsc{Journal}:\;\;\textit{Frontiers in Artificial Intelligence}, Volume 3 \\
\textsc{DOI}:\;\;10.3389/frai.2020.00033 \\
\textsc{Note}:\;\;The appendix has been included as section~\ref{sec:tuning-appendix}.

\section{Abstract}

@@

\section{Target labels for tuning group fairness}
We will start by describing several notions of group fairness.
For each individual, we have a vector of non-sensitive attributes $x\in\mathcal{X}$, a class label $y\in\mathcal{Y}$, and a sensitive attribute $s\in\mathcal{S}$ (e.g.\ racial origin or gender).
%
We focus on the case where $s$ and $y$ are binary. 
%
We assume that a positive label $y=1$ corresponds to a positive outcome for an individual -- for example, being accepted for a loan.
%
%
\emph{Group fairness} balances a certain condition between groups of individuals with different sensitive attribute, $s$ versus $s\sp{\prime}$. 
%
The term $\hat{y}$ below is the prediction of a machine learning model that,
in most works, uses only non-sensitive attributes $x$.
%
Several group fairness criteria have been proposed (e.g. \cite{zafar2017fairnesstreatment,chouldechova2017fair,hardt2016equality}):
\begin{align}
& \text{equality of positive rate (Demographic Parity):}\nonumber\\
& \text{Pr}(\hat{y}=1|s)=\text{Pr}(\hat{y}=1|s\sp{\prime}) \label{eq:eq_dem_par}\\
& \text{equality of accuracy:}\nonumber\\
& \text{Pr}(\hat{y}=y|s)=\text{Pr}(\hat{y}=y|s\sp{\prime}) \\
& \text{equality of true positive rate (Equality of Opportunity):} \nonumber\\
& \text{Pr}(\hat{y}=1|s,y=1)=\text{Pr}(\hat{y}=1|s\sp{\prime},y=1)  \label{eq:eq_opp_cri}~.
\end{align}
\emph{Equalised odds} criterion corresponds to Equality of Opportunity (\ref{eq:eq_opp_cri}) plus equality of false positive rate.

@@

\subsection{Choice of parameters}\label{sec:dp}
A balanced dataset is characterised by an independence of the label $\bar{y}$ and the sensitive attribute $s$.
%As our model is predicting $\bar{y}$, we want this independence to hold for $\bar{y}$.
%The graphical model in Fig~\ref{fig:biasfeaturebiaslabel} does not imply $\bar{y} \bot s$,
Given that we have complete control over the \emph{transition probabilities},
we can ensure this independence by requiring $P(\bar{y}=1|s=0)=P(\bar{y}=1|s=1)$.
Our constraint is then that both of these probabilities are equal to the same value,
which we will call the target rate $\mathit{PR}_t$ (``PR'' as \emph{positive rate}):
\begin{align}
  &P(\bar{y}=1|s=0) \overset{!}{=} \mathit{PR}_t\quad\text{and}%\nonumber\\
  \quad P(\bar{y}=1|s=1) \overset{!}{=} \mathit{PR}_t~.
\end{align}
This leads us to the following constraints for $s\prime\in\{0, 1\}$:
\begin{align}
  \mathit{PR}_t &= P(\bar{y}=1|s=s\prime) %\nonumber\\
  =\sum\limits_y P(\bar{y}=1|y,s=s\prime)\, P(y|s=s\prime).\label{eq:dpconstraint}
\end{align}
%$P(y|s=s\prime)$ has to be set to the value at which we want our constraint to hold.
We call $P(y=1|s=j)$ the base rate $\mathit{PR}_b^j$ which we estimate from the training set:
\begin{align}
  P(y=1|s=i) = \frac{\text{number of points with } y=1 \text{ in group }i}
  {\text{number of points in group }i}~.
\end{align}
Expanding the sum, we get
\begin{align}
  \mathit{PR}_t %\nonumber\\
  =\,\, P(\bar{y}=1|y=0,s=s\prime) \cdot (1-\mathit{PR}_b^1) 
  +P(\bar{y}=1|y=1,s=s\prime) \cdot \mathit{PR}_b^1~.\label{eq:dpconstexpanded}
\end{align}
This is a system of linear equations consisting of two equations (one for each value of $s\prime$)
and four free variables: $P(\bar{y}=1|y,s)$ with $y,s\in\{0, 1\}$.
The two unconstrained degrees of freedom determine how strongly the accuracy will be affected by the fairness constraint.
If we set $P(\bar{y}=1|y=1,s)$ to 0.5,
then this expresses the fact that a train label $y$ of $1$ only implies a target label $\bar{y}$ of $1$ in 50\% of the cases.
In order to minimise the effect on accuracy,
we make $P(\bar{y}=1|y=1,s)$ as high as possible and $P(\bar{y}=1|y=0,s)$, conversely, as low as possible.
However, the lowest and highest possible values are not always 0 and 1 respectively.
To see this, we solve for $P(\bar{y}=1|y=0,s=j)$ in \eqref{eq:dpconstexpanded}:
\begin{align}
  &P(\bar{y}=1|y=0,s=j) %\nonumber\\
  =\,\, \frac{\mathit{PR}_b^j}{1-\mathit{PR}_b^j} \left(\frac{\mathit{PR}_t}{\mathit{PR}_b^j} - P(\bar{y}=1|y=1,s=j)\right)~.
\end{align}
If $\nicefrac{\mathit{PR}_t}{\mathit{PR}_b^j}$ were greater than 1,
then setting $P(\bar{y}=1|y=0,s=j)$ to 0
would imply a $P(\bar{y}=1|y=1,s=j)$ value greater than 1.
A visualisation that shows why this happens can be found in the Supplementary Material.
We thus arrive at the following definitions:
\begin{align}
  P(\bar{y}=1|y=1,s=j)&=\begin{cases}
    1 &\text{if }\mathit{PR}_t>\mathit{PR}_b^j\\
    \frac{\mathit{PR}_t}{\mathit{PR}_b^j} &\text{otherwise.}
  \end{cases}%
  \label{eq:tau-11}\\
  P(\bar{y}=1|y=0,s=j)&=\begin{cases}
    \frac{\mathit{PR}_t-\mathit{PR}_b^j}{1-\mathit{PR}_b^j} &\text{if }\mathit{PR}_t>\mathit{PR}_b^j\\
    0 &\text{otherwise.}
  \end{cases}%
  \label{eq:tau-01}
\end{align}
Algorithm~\ref{alg:parity} shows pseudocode of the procedure, including the computation of the allowed minimal and maximal value.

@@

\subsection{Conditionally balanced dataset}
\noindent There is a fairness definition related to demographic parity which allows conditioning on ``legitimate'' risk factors $\ell$
when considering how equal the demographic groups are treated~\citep{corbett2017algorithmic}.
This cleanly translates into balanced datasets which are balanced conditioned on $\ell$:
\begin{align}
  &P(\bar{y}=1|\ell=\ell\prime, s=0) \overset{!}{=} P(\bar{y}=1|\ell=\ell\prime, s=1)~.
\end{align}
We can interpret this as splitting the data into partitions based on the value of $\ell$,
where the goal is to have all these partitions be balanced.
This can easily be achieved by our method by setting a $\mathit{PR}_t(\ell)$ for each value of $\ell$
and computing the transition probabilities for each sample depending on $\ell$.

\section{Related work}\label{sec:relatedwork}
There are several ways to enforce fairness in machine learning models:
as a pre-processing step
% (e.g.~),
\citep{kamiran2012data,louizos2016variational,lum2016statistical,zemel2013learning,Chiappa19,QuaShaTho19},
% kusner2017counterfactual
as a post-processing step \citep{feldman2015certifying,hardt2016equality}, %(e.g.~\citep{feldman2015certifying,hardt2016equality}),
 or as a constraint during the learning phase
% (e.g.~).
\citep{calders2009building,zafar2017fairnesstreatment,zafar2017fairnessconstraints,DonOneBenShaetal18,DimLiuParRad19}.
%
Our method enforces fairness during the learning phase (an in-processing approach) but, unlike other approaches, we do not cast fair-learning as a \emph{constrained} optimisation problem.
Constrained optimisation requires a customised procedure.
%
In \citet{GohCotGupFri16}, \citet{zafar2017fairnessconstraints}, and \citet{zafar2017fairnesstreatment},
% In~\citet{GohCotGupFri16,zafar2017fairnesstreatment,zafar2017fairnessconstraints},
suitable majorisation-minimisation/convex-concave procedures~\citep{LanSri09} were derived.
%
Furthermore, such constrained optimisation approaches may lead to more unstable training,
and often yield classifiers with both worse accuracy and more unfair~\citep{CotJiaWanNar18}.

@@

The massaging approach uses a classifier to first rank all samples according to their probability of having a positive label ($y=1$)
and then flips the labels that are closest to the decision boundary such that the data then satisfies demographic parity.
This \emph{pre-processing} approach is similar in spirit to our \emph{in-processing} method but differs in the execution.
In our method (section \ref{sec:dp}), ``ranking'' and classification happen in one step and labels are not explicitly flipped
but assigned probabilities of being flipped.
%{\color{red}This protects us from flipping the wrong labels.}

@@

In addition to the GP and LR baselines, we compare our proposed model with the following methods:
Support Vector Machine (\emph{SVM}), \emph{Kamiran \& Calders}~\citep{kamiran2012data} (``reweighing'' method),
\emph{Agarwal et al.}~\citep{AgaBeyDudLanetal18} (using logistic regression as the classifier)
% Logistic Regression (\emph{LR})
and several methods given by \citet{zafar2017fairnessconstraints,zafar2017fairnesstreatment},
%\citet{zafar2017fairnessconstraints} and \citet{zafar2017fairnesstreatment},
which include maximising accuracy under demographic parity fairness constraints (\emph{ZafarFairness}),
maximising demographic parity fairness under accuracy constraints (\emph{ZafarAccuracy}),
and removing disparate mistreatment by constraining the false negative rate (\emph{ZafarEqOpp}).
% We make use of the fairness comparison by \citet{friedler2018comparative} %\citet{friedler2018comparative},
Every method is evaluated over 10 repeats that each have different splits of the training and test set.

\subsection{Results for Demographic Parity on Adult dataset}\label{sssec:demparresults}
\begin{figure}[t]
  \begin{center}
    \centerline{\includegraphics[width=0.98\columnwidth]{./figures/adult_tar_choice.eps}}
    \caption{Accuracy and fairness (demographic parity) for various target choices.
      (a): Adult dataset using race as the sensitive attribute;
      (b): Adult dataset using gender.
      Centre of the cross is the mean; height and width of the box encode half of standard derivation of accuracy and disparate impact.
    }%
    \label{fig:tar}
  \end{center}
  % \vspace*{-0.8cm}
% \caption{Accuracy and fairness for various targets.  w.r.t. accuracy and width of the box encodes standard derivation w.r.t. fairness criterion}
\end{figure}%
%
Following \citet{zafar2017fairnessconstraints} we evaluate demographic parity on the Adult dataset.
Table~\ref{tab:dempar} shows the accuracy and fairness for several algorithms.
In the table, and in the following, we use $\mathit{PR}_{s=i}$ to denote
the observed rate of positive predictions per demographic group $P (\hat{y} = 1 | s =i)$.
Thus, $\mathit{PR}_{s=0} / \mathit{PR}_{s=1}$ is a measure for demographic parity,
where a completely fair model would attain a value of $1.0$.
This measure for demographic parity is also called ``disparate impact''
(see e.g.~\cite{feldman2015certifying,zafar2017fairnesstreatment}).
%
As the results in Table~\ref{tab:dempar} show, FairGP and FairLR are clearly fairer than the baseline GP and LR\@.
We use the mean ($\mathit{PR}_t^{avg}$) for the target acceptance rate.
The difference between fair models and unconstrained models is not as large with \emph{race} as the sensitive attribute,
as the unconstrained models are already quite fair there.
The results of FairGP are characterised by high fairness and high accuracy.
FairLR achieves similar results to FairGP, but with generally slightly lower accuracy but better fairness.
We used the two step procedure of \citet{DonOneBenShaetal18}
to verify that we cannot achieve the same fairness result with just parameter search on LR\@.

In Fig.~\ref{fig:tar}, we investigate which choice of target ($\mathit{PR}_t^{avg}$, $\mathit{PR}_t^{min}$ or $\mathit{PR}_t^{max}$)
gives the best result.
We use $\mathit{PR}_t^{avg}$ for all following experiments as this is the fairest choice (cf. Section~\ref{sec:dp}).
The Fig.\ref{fig:tar}(a) shows results from Adult dataset with \emph{race} as sensitive attribute
where we have $\mathit{PR}_t^{min}=0.156$, $\mathit{PR}_t^{max}=0.267$ and $\mathit{PR}_t^{avg} =0.211$.
$\mathit{PR}_t^{avg}$ performs best in term of the trade-off.%equal (for \emph{race} as sensitive attribute)
%or better (for \emph{gender}) compared with the two other possibilities.
% The variant that uses $s$ for training and prediction (``use $s$'') performs significantly better here.
%\emph{ZafarAccuracy} can achieve good fairness results at the cost of accuracy.
\begin{figure}[t]
  \centering
  \includegraphics[width=0.98\textwidth]{./figures/adult_parity_scatter_pr_pr.eps}
  \caption{
    Predictions with different target acceptance rates (demographic parity) for 10 repeats.
    (a): $\mathit{PR}_{s=0}$ vs $\mathit{PR}_{s=1}$ using race as the sensitive attribute;
    (b): $\mathit{PR}_{s=0}$ vs $\mathit{PR}_{s=1}$ using gender.
  }%
  \label{fig:adult_parity_scatter_pr_pr}
\end{figure}%
\begin{figure}[t]
  \centering
  \includegraphics[width=0.98\textwidth]{./figures/adult_parity_scatter_acc.eps}
  \caption{
    Predictions with different target acceptance rates (demographic parity) for 10 repeats.
    (a): disparate impact vs accuracy on Adult dataset using race as the sensitive attribute;
    (b): disparate impact vs accuracy using gender.
  }%
  \label{fig:adult_parity_scatter_acc}
\end{figure}%

@@

\subsection{Results for Equality of Opportunity on ProPublica dataset.}\label{sssec:eqoppresults}
%
For equality of opportunity,
we again follow \citet{zafar2017fairnesstreatment} and evaluate the algorithm on the ProPublica dataset.
As we did for demographic parity,
we define a measure of equality of opportunity via the ratio of the true positive rates (TPRs) within the demographic groups.
We use $\mathit{TPR}_{s=i}$ to denote the observed TPR in group $i$: $P(\hat{y}=1|y=1, s=i)$,
and $\mathit{TNR}_{s=i}$ for the observed true negative rate (TNR) in the same manner.
The measure is then given by $\mathit{TPR}_{s=0} / \mathit{TPR}_{s=1}$.
A perfectly fair algorithm would achieve $1.0$ on the measure.%
\begin{figure*}[t]
  \centering
  \includegraphics[width=0.98\textwidth]{./figures/propublica_opp_box_with_agarwal.eps}
%   \includegraphics[width=0.45\textwidth]{propublica-recidivism_race_opp_box.eps}
%   \includegraphics[width=0.45\textwidth]{propublica-recidivism_sex_opp_box.eps}
  \caption{Accuracy and fairness (with respect to \emph{equality of opportunity}) for various methods on ProPublica dataset.
    (a): using race as the sensitive attribute; (b): using gender.
    A completely fair model would achieve a value of 1.0 in the x-axis.
    %The best value for the TPR ratio is $1.0$.
    See Fig.~\ref{fig:propublica_opp_scatter_tpr}(a) and (b) on how these choices of PR setting translate to
    $\mathit{TPR}_{s=0}$ vs $\mathit{TPR}_{s=1}$. %and TPR vs TNR.
  }%
  \label{fig:propublica_opp_box}
\end{figure*}%
\begin{figure*}[!ht]
  \centering
  \includegraphics[width=0.98\textwidth]{./figures/propublica_opp_scatter_tpr.eps}
  \caption{%
    Fairness measure $\mathit{TPR}_{s=0}$ vs $\mathit{TPR}_{s=1}$ (\emph{equality of opportunity}) for different target PRs ($\mathit{PR}_t$).
    (a): on dataset ProPublica recidivism using race as the sensitive attribute;
    (b): using gender.
  }%
  \label{fig:propublica_opp_scatter_tpr}
\end{figure*}%

@@

\section{Discussion and conclusion}
Fairness is fundamentally not a challenge of algorithms alone, but very much a sociological challenge.
%
A lot of proposals have emerged recently for defining and obtaining fairness in machine learning-based decision making systems. 
%
% A vast amount
The vast majority of academic work has focused on two categories of definitions: statistical (group) notions of fairness and individual notions of fairness (see~\citet{verma2018fairness} for at least twenty different notions of fairness).
%
Statistical notions are easy to verify but do not provide protections to individuals. 
%
Individual notions do give individual protections but need strong assumptions, such as the availability of an agreed-upon similarity metric, which can be difficult in practice.
%
We acknowledge that a proper solution to algorithmic fairness cannot rely on statistics alone.
%
Nevertheless, these statistical fairness definitions can be helpful in understanding the problem and working towards solutions.
%
To facilitate this, at every step, the trade-offs that are present should be made very clear
and long-term effects have to be considered as well~\citep{liu2018delayed,kallus2018residual}.
% For example, it can be the case that false positives and false negatives are not equally costly,
% in which case a closer inspection of
